Effects of Trauma Life Support Training Programs on Patient Outcomes: A Pilot Cluster Randomised Trial

### 2024-09-26 pilot-publication/manuscript.Rmd
Revise manuscript

After training, a total of `r m30d.standard.care` patients in the standard care arm died within 30 days, compared to `r m30d.atls` patients in the ATLS^®^ arm and `r m30d.ptc` patients in the PTC arm. The corresponding figures for in-hospital mortality were `r inhosp.standard.care`%, `r inhosp.atls`%, and `r inhosp.ptc`%. Absolute and relative changes from baseline, as well as absolute and relative risk differences, are shown in Table \@ref(tab:outcomes). Overall, mortality was lower in the ATLS^®^ and PTC arms compared to the standard care arm. The results for all other outcomes are shown in Supplementary Materials.

```{r outcomes, echo = FALSE}
outcomes.table
```

@@

We found that the ATLS^®^ arm had lower 30-day mortality compared to the
PTC and standard care arms. We also found that the PTC arm had lower mortality than the standard care arm. These findings indicate a large potential effect of training physicians in trauma life support, but it is important to note that this pilot study was not powered to detect any differences in outcomes. The arms differed considerably in sample size, with the ATLS^®^ arm having the smallest sample size. This difference resulted from the randomisation process with a small number of heterogeneous clusters.

The primary patient participant outcome, all-cause 30-day mortality,
was missing in `r rate.lost.to.follow.up`% of patients. This may
appear high, especially compared to for example the CRASH-2 and
REACT-2 trials, which report missing primary outcome in less than
0.01% of patients [@Roberts2013; @Sierink2016]. Like many other trauma
trials, both CRASH-2 and REACT-2 used in-hospital mortality as their
primary outcome measure, whereas we attempted to follow up patients
after discharge. Our missing data rate for in-hospital mortality was
only `r p.missing.in.hospital.mortality`%, comparable to previous trials.

@@

During the course of this pilot we deviated from the protocol in
multiple ways, and provide a detailed list as Supplementary
material. Some key limitations of this pilot and therefore lessons to
be learned and factored into the design of the full-scale trial
include the lower than expected enrolment rates of some centres,
centre specific management routines, and difficulties in collecting
data on complications and cause of death. We minimised the impact of
the lower than expected enrolment rates by including a seventh centre,
but on-site observations of patient volumes are likely to be needed
for the full-scale trial. We decided to be pragmatic in selecting
which residents to train and how to structure the data collection
depending on how and by whom patients were initially managed, but this
flexibility will need to be built into the full-scale trial
protocol. Finally, we found that data on complications and cause of
death were hard to identify and therefore the full-scale trial will
need to include longer training of research officers.

@@

\pagebreak
# Supplementary material

>>>>

After training, a total of `r m30d.standard.care` patients in the standard care arm died within 30 days, compared to `r m30d.atls` patients in the ATLS^®^ arm and `r m30d.ptc` patients in the PTC arm. The corresponding figures for in-hospital mortality were `r inhosp.standard.care`%, `r inhosp.atls`%, and `r inhosp.ptc`%. Absolute and relative changes from baseline, as well as absolute and relative risk differences, are shown in Table \@ref(tab:outcomes). Overall, both in-hospital and 30-day mortality were substantially lower in the ATLS^®^ and PTC arms compared to the standard care arm, but the absolute numbers of deaths in the ATLS and PTC arms were very small. The results for all other outcomes are shown in Supplementary Materials.

```{r outcomes, echo = FALSE}
outcomes.table.with.differences
```

@@

We found that the ATLS^®^ arm had lower 30-day mortality compared to the PTC and standard care arms. We also found that the PTC arm had lower mortality than the standard care arm. These findings indicate a large potential effect of training physicians in trauma life support, but it is important to note that this pilot study was not powered to detect any differences in outcomes. The arms differed considerably in sample size, with the ATLS^®^ arm having the smallest sample size. This difference most likely resulted from the randomisation process with a small number of heterogeneous clusters.

All-cause 30-day mortality was missing in `r rate.lost.to.follow.up`% of patients. This may appear high, especially compared to for example the CRASH-2 and REACT-2 trials, which report missing primary outcome in less than 0.01% of patients [@Roberts2013; @Sierink2016]. Like many other trauma trials, both CRASH-2 and REACT-2 used in-hospital mortality as their
primary outcome measure, whereas we attempted to follow up patients after discharge. Our missing data rate for in-hospital mortality was only `r p.missing.in.hospital.mortality`%, comparable to previous trials.

@@

During the course of this pilot we deviated from the protocol in several ways, and provide a detailed list as Supplementary
material. Some key limitations of this pilot and therefore lessons to be learned and factored into the design of the full-scale trial include the lower than expected enrolment rates of some centres, centre specific management routines, and difficulties in collecting data on complications and cause of death.

We minimised the impact of the lower than expected enrolment rates by including a seventh centre, but on-site observations of patient volumes are likely to be needed for the full-scale trial. We decided to be pragmatic in selecting which residents to train and how to structure the data collection
depending on how and by whom patients were initially managed, but this flexibility will need to be built into the full-scale trial protocol. Finally, we found that data on complications and cause of death were hard to identify and therefore the full-scale trial will need to include longer training of research officers.

@@

\pagebreak

# Supplementary material
